Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-12 manuscript/manuscript.tex
more theory section stuff

When mutations are weakly deleterious, their frequency trajectories are no
longer well-approximated by deterministic models. This is because the selection
on an allele is only weak relative to the scale of stochastic perturbations it
encounters along its trajectory. When genetic drift is the only source of
randomness in allele frequency change, the relevant scale of stochastic
perturbations is the drift-effective population size $N$ and selection dynamics
are governed by compound parameter $2Ns$. When $2Ns \gg 1$, the stochastic
component of allele frequency change is minuscule relative to selection and can
be ignored as the classic BGS theory does; this is why the rescaling factor
under the BGS model does not depend on $N$. However, when $2Ns \le 1$
stochastic perturbations are large relative to selection, and a weakly
deleterious allele can drift up to intermediate frequencies before eventual
loss or fixation. Weakly deleterious alleles can persist in the population much
longer, and consequently

When this occurs, the distribution of the number of deleterious mutations (and
its corresponding fitness distribution) are no longer stationary and become a
traveling wave \parencite{Rouzine2008-qz,Good2013-lp}. Furthermore, individuals
in the least-loaded class are stochastically lost, fixing those mutations with
a click of ``Muller's ratchet" \parencite{Muller1964-ki}. Determining both the
rate of Muller's ratchet \parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz}
and the shape of genealogies under weak selection
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc} have been stubborn
problems

>>>>

When mutations are weakly deleterious, their allele frequencies are strongly
influenced by stochastic perturbations and their frequency trajectories are no
longer well-approximated by deterministic models. When genetic drift is the
only source of randomness in allele frequency change, the relevant scale of
stochastic perturbations is determined by the drift-effective population size
$N_d$ (e.g. due to only non-selective process). For reasons that will soon be
clear, we consider a general expression of the drift-effective population size
that allows for extra non-heritable variability in offspring number $V_k$,

\begin{align}
    N_d = \frac{2N}{V_k + 2}
\end{align}

where if $V_k = 2$, this reduces to the classic Wright--Fisher model
\parencite{Wright1938-tv}.

The insight of nearly neutral theory (XXX) was that the qualitative behavior of
selected sites is determined by the compound parameter $2N_d s$. When $2N_d s
\gg 1$, the stochastic component of allele frequency change is minuscule
relative to selection and can be ignored; this is why the factor that rescales
the drift-effective population size under the BGS model does not depend on
$N_d$. However, when $2N_d s \le 1$ stochastic perturbations are on a scale
equal to or greater than the changes due to selection. In this regime, weakly
deleterious alleles can drift up to intermediate frequencies before their
eventual loss or fixation. When this occurs, classic BGS theory breaks down for
several reasons. First, the ``delay phase" is no longer negligible as lineages
carrying weakly deleterious mutations can persist on coalescent timescales.
This distorts genealogies away those expected under neutrality
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc}. Second, the
distribution of the number of deleterious mutations (and its corresponding
fitness distribution) are no longer stationary and become a traveling wave
\parencite{Rouzine2008-qz,Good2013-lp,Gessler1995-hz} towards reduced
population fitness (since beneficial mutations are ignored). Consequently,
individuals in the least-loaded class are stochastically lost, fixing those
mutations with a click of ``Muller's ratchet" \parencite{Muller1964-ki}.
Determining both the rate of Muller's ratchet
\parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz} and the shape of
genealogies under weak selection have been stubborn open problems in
evolutionary genetics.

To further complicate matters, both strong and weakly deleterious mutations
create \emph{heritable} fitness variation. In turn, the presence of heritable
fitness variation generates another source of randomness in allele frequency
change known as genetic \emph{draft} \parencite{Neher2013-dz}. Genetic draft
occurs when alleles become randomly associated with different fitness
backgrounds each generation, which stochastically perturbs their trajectories.
Like genetic drift, perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, the stochastic
perturbations each generation due to draft are autocorrelated as long as
segregation and recombination do not decouple an allele from its fitness
background and the selective environment does not change
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. When selection is
strong,

When selection is
weak and fitness is polygenic, the stochasticity introduced by draft is
well-approximated by rescaling the drift-effective population size

that exist when there are heritable differences among haplotypes.
Mathematical models of genetic draft

Gillespie2000-mh

,Robertson1961-ho,Santiago1995-hx,Santiago1998-bs,Barton2000-zg
